Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-28 paper.tex
Some updates to first few sections.

One of the key reasons for \msprime's substantial performance advantage
over other simulators~\citep{kelleher2016efficient}
is its usage of the ``succinct tree sequence''
 data stucture to represent simulation results.
The succinct tree sequence (usually abbreviated to ``tree sequence'')
is a data structure for concisely encoding
genetic ancestry and sequence variation.
The basic concepts were introduced by \cite{kelleher2016efficient}
and originally developed as part of \msprime. We subsequently
extracted the tree sequence code that was independent of simulations
from \msprime\ into \tskit,
a standalone Python and C library~\citep{tskit2021tskit}.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarise
some aspects that are important for simulation here.

% Encoding ancestry
Given a set of sample chromosomes, the topology
of their genetic ancestry is described by a set of \emph{nodes}
and \emph{edges} (Fig.~\ref{fig-ts-example}).
A node in tree sequence terminology corresponds
to a single chromosome strand, associated with its birth time
and other optional metadata. A node may be
associated with an \emph{individual}, so that, for example, a diploid
individual is associated with two separate nodes (we omit
the individual table in Fig.~\ref{fig-ts-example} as it is optional).
Nodes define haplotypes that exist at a specific time within
some individual, and edges define genetic inheritence
relationships between these nodes. An \emph{edge} consists of a
parent node, a child node and the left and right coordinates
of the contiguous genomic interval over which
the child inherited genetic material from the parent.
Critically, edges can span multiple marginal trees: for
example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral chromosomes and recording the
genomic intervals over which parent/child relationships exist,
generalises the original
``coalescence records'' concept~\citep{kelleher2016efficient},
and is the key to the efficiency of tree
sequences~\citep{
kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.

@@

% OK, back to tree sequences: what about sequence variation?
The final output of most population genetic simulations is some representation
of sequence variation among the specified samples. For coalescent
simulations, we usually have three steps: (1) simulate
the genetic ancestry, and optionally output the resulting marginal trees;
(2) simulate sequence evolution conditioned on this ancestry by generating
mutations (see the \nameref{sec-mutations} section); and (3) output the
resulting sequences by percolating the effects of the mutations through
the trees. Information about the mutations
themselves---e.g., where they have occurred on the
trees---is usually not retained or made available for subsequent analysis.
In \msprime, however, we skip step (3), instead using \tskit's
data model of ancestry and mutations
to represent the simulated sequences.
As illustrated in Fig.~\ref{fig-ts-example}, mutations are a
fully integrated part of \tskit's tree sequence data model, and genetic
variation is encoded by recording sites at which mutations
have occurred, and where each mutation at those sites
has occurred on the marginal tree.
Crucially, the sequences themselves are never stored, or indeed directly
represented in memory (unless explicitly decoded using \tskit\ APIs).
It may at first seem inconvenient to have only this indirect representation
of the genome sequences, but it is extremely powerful.
Firstly, the storage space required for simulations is dramatically
reduced. For a simulation of $n$ samples with $m$ variant sites,
we would require $O(nm)$ space to store the sequence data as a variant
matrix. However, if this simulation was of a recombining genome
with $t$ trees, then the \tskit\ tree sequence encoding
requires $O(n + t + m)$ space,
assuming we have $O(1)$ mutations at each site~\citep{kelleher2016efficient}.
For large sample sizes, this difference is profound, making it
conceivable, for example, to store the genetic ancestry
and variation data for the entire human population on a
laptop~\citep{kelleher2019inferring}. Besides the huge
difference in storage efficiency, it is also often far more efficient
to compute statistics of the sequence data from the
trees and mutations than it is to work with the sequences themselves.
For example, computing Tajima's $D$ from simulated data using \tskit\
is several orders of magnitude faster than efficient variant matrix
libraries for large sample sizes~\citep{ralph2020efficiently}.

@@

@@

The more recent trend has been to move away from this file and command-line
driven approach and to instead provide direct interfaces to the simulation
engines via an Application Programming Interface (API)~\citep[e.g.][]{
thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
The primary interface for \msprime\ is through a thoroughly documented and
stable Python
API, which has allowed a number of downstream tools to be built on top of
% Found these through the Dependent packages on GitHub
it~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
adrion2020community,adrion2020predicting, kamm2020efficiently,
mckenzie2020ipcoal, montinaro2020revisiting,
de2021geonomics,rivera2021simulation}.
As well as providing a stable and efficient platform for building
downstream applications, \msprime's Python API makes it much easier to
build reproducible simulation pipelines, as the entire workflow can
be encapsulated in a single script, and package and version
dependencies can be explicitly stated using the \texttt{pip}
or \texttt{conda} package managers.
For example, the errors made~\citep{ragsdale2020lessons,martin2020erratum}
in the influential simulation analysis of
\cite{martin2017human} were only detected because the pipeline
could be easily run and reanalysed.

A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
designed in part to avoid sources of error (see the Demography section) but
also to provide more appropriate defaults while keeping compatibility with
existing code. In the new APIs, ancestry and mutation simulation are fully
separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
functions replacing the legacy \texttt{simulate} function. Among other changes,
the new APIs default to discrete genome coordinates and finite sites mutations,
resolving a major source of confusion and error. The 0.x APIs are fully
supported and tested, and will be maintained for the foreseeable future.
We also provide an \ms\ compatible
command line interface to support existing workflows, and the
\texttt{msp} program that supports specifying demographic models in
Demes format~\citep{gower2021demes} from the command line.

% The output - what do we get out?
\subsection*{Data analysis}
\label{sec-data-analysis}
The point of running simulations is to generate data, and there
are two important ways in which the \tskit\ tree sequences output by
\msprime\ are more useful than established methods: (1) tree
sequences are more efficient; and (2) tree sequences contain
more information. We will consider these points in turn.

@@

Classical text based output formats like \ms\ are inefficient to process,
but also lack a great deal of important information.
The tree-by-tree topology information output by simulators in Newick
format lacks any concept of node (in the tree sequence sense) identity,
and means that we cannot reliably infer information about ancestors
from the output. Because Newick stores branch lengths rather
than node times, numerical precision issues also arise when summing
over many short branches~\citep{mcgill2013graphml}.
% JK: not sure this is worth leaving in tbh
This lack of reliable information about ancestors in simulation data
has lead, for example, to the definition of segments of
identity by descent as a set of samples having the same time
to their most recent common ancestor along some section of the
genome~\citep{yang2016rapidly}.
Numerous forks
of simulators have been created to access information not provided
in the output. For example, \ms\  has been forked to
output information about migrating segments~\citep{rosenzweig2016powerful},
information about ancestral lineages~\citep{chen2013asymptotic},
an array of population genetic statistics~\citep{ramos2007mlcoalsim}
% Any more?
and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
has in turn been forked to output information on local
ancestry~\citep{racimo2017archaic}.
All of this information
is either directly available by default in \msprime, or can be optionally
stored via options such as \texttt{record\_migrations} or
\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
can be efficiently and conveniently processed via \tskit\ APIs.

@@

Part of the reason for this poor record of software reuse and modularity is the
lack of standardised file formats, and in particular, the lack of common
library infrastructure to abstract the details of interchanging simulation
data. On the surface, it would appear that \msprime\ is also guilty of this
lack of modularity since it supports simulating both ancestry and mutations.
However, this is essentially a historical accident and because of the \tskit\
library, the mutation process in \msprime\ could just as easily
be performed in a separate package.
The mature and efficient C and Python interfaces for
\tskit\ makes it straightforward to add further information to
an existing file, and because of its efficient data interchange mechanisms,
there is no performance penalty for performing additional operations
in a different Python package.
Thanks to this interoperability, \msprime's
mutation generator can work with \emph{any} \tskit\ tree sequence,
be it simulated using \SLiM~\citep{haller2019slim} or
\fwdpy~\citep{thornton2014cpp}, or estimated from real data
using~\texttt{tsinfer}/\texttt{tsdate}~\citep{kelleher2019inferring,wohns2021unified}
or \texttt{Relate}~\citep{speidel2019method}. It is a modular
component intended to fit into a larger software ecosystem, and
is in no way dependent on \msprime's ancestry simulator.

As well as providing a new API (see the \nameref{sec-sim-interface} section)
which emphasises this logical split between ancestry and mutation simulation,
we have greatly extended the sophistication of
\msprime's mutation generation engine for version 1.0,
achieving near feature-parity with \SeqGen.
We  support a large number of mutation models, including the
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
A general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
can be easily and efficiently defined in user code.
Mutation rates can vary along the genome, and multiple mutation
models can be imposed on a tree sequence by overlaying mutations
in multiple passes.
We have extensively validated the results of mutation simulations
against both theoretical expectations and output from
\SeqGen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

Simulating mutations in \msprime\ is efficient.
Fig.~\ref{fig-mutations-perf} shows the time required to generate
mutations (using the default JC69 model) on
simulated tree sequences for a variety of mutation
rates as we vary the number of samples
(Fig.~\ref{fig-mutations-perf}A) and the sequence
length (Fig.~\ref{fig-mutations-perf}B).
For example, the longest running simulation in
Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
generate an average of 1.5 million mutations over 137,081 trees
in a tree sequence with 508,125 edges.
This efficiency for large numbers of trees is possible because
the tree sequence encoding allows us to generate mutations
on an edge-by-edge basis
(see Fig.~\ref{fig-ts-example} and Appendix~\ref{app-mutation-algorithm}),
rather than tree-by-tree and branch-by-branch as would otherwise be required.
In the above example from Fig.~\ref{fig-mutations-perf}B,
we would have to consider 273,887,838 branches
since there are 137,081 trees and 1,998 branches in each
tree ($\sim500$ times more work).
% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
Even if we have a tree sequence consisting of a single tree
(so that there is no advantage to working edge-by-edge),
\msprime's mutation generator is still very efficient.
For example, we simulated mutations under the BLOSUM62 amino
acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim$260,000 mutations) in about $0.8$ seconds, including
the time required for file input and output.
We do not attempt a systematic benchmarking of \msprime's
mutation generation code against other methods because at this scale it is
difficult to disentangle the effects of inefficient input and
output formats from the mutation generation algorithms.
It seems quite unlikely, however,
that generating mutations with \msprime\ would be a bottleneck in any
realistic analysis.

>>>>

One of the key reasons for \msprime's substantial performance advantage
over other simulators~\citep{kelleher2016efficient}
is its usage of the ``succinct tree sequence''
 data stucture to represent simulation results.
The succinct tree sequence (usually abbreviated to ``tree sequence'')
is a data structure for concisely encoding
genetic ancestry and sequence variation.
The basic concepts were introduced by \cite{kelleher2016efficient}
and originally developed as part of \msprime. We subsequently
extracted the tree sequence code that was independent of simulations
from \msprime\ into
a standalone Python and C library, called \tskit~\citep{tskit2021tskit}.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarise
some aspects that are important for simulation here.

% Encoding ancestry
Given a set of sample chromosomes, the topology
of their genetic ancestry is described by a set of \emph{nodes}
and \emph{edges} (Fig.~\ref{fig-ts-example}).
A node in tree sequence terminology corresponds
to a single chromosome strand, associated with its birth time
and other optional information. A node may be
associated with an \emph{individual}, so that, for example, a diploid
individual is associated with two separate nodes (we omit
the individual table in Fig.~\ref{fig-ts-example} as it is optional).
Nodes define haplotypes that exist at a specific time within
some individual, and edges define genetic inheritence
relationships between these nodes. An \emph{edge} consists of a
parent node, a child node and the left and right coordinates
of the contiguous genomic interval over which
the child inherited genetic material from the parent.
Critically, edges can span multiple marginal trees: for
example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral chromosomes and recording the
genomic intervals over which parent/child relationships exist,
generalises the original
``coalescence records'' concept~\citep{kelleher2016efficient},
and is the key to the efficiency of tree
sequences~\citep{
kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.

@@

% OK, back to tree sequences: what about sequence variation?
The final output of most population genetic simulations is some representation
of sequence variation among the specified samples. For coalescent
simulations, we usually have three steps: (1) simulate
the genetic ancestry, and optionally output the resulting marginal trees;
(2) simulate sequence evolution conditioned on this ancestry by generating
mutations (see the \nameref{sec-mutations} section); and (3) output the
resulting sequences by percolating the effects of the mutations through
the trees. Information about the mutations
themselves---e.g., where they have occurred on the
trees---is usually not retained or made available for subsequent analysis.
In \msprime, however, we skip step (3), instead using \tskit's
combined data model of ancestry and mutations
to represent the simulated sequences.
As illustrated in Fig.~\ref{fig-ts-example}, mutations are a
fully integrated part of \tskit's tree sequence data model, and genetic
variation is encoded by recording sites at which mutations
have occurred, and where each mutation at those sites
has occurred on the marginal tree.
Crucially, the sequences themselves are never stored, or indeed directly
represented in memory (unless explicitly decoded using \tskit\ APIs).
It may at first seem inconvenient to have only this indirect representation
of the genome sequences, but it is extremely powerful.
Firstly, the storage space required for simulations is dramatically
reduced. For a simulation of $n$ samples with $m$ variant sites,
we would require $O(nm)$ space to store the sequence data as a variant
matrix. However, if this simulation was of a recombining genome
with $t$ trees, then the \tskit\ tree sequence encoding
requires $O(n + t + m)$ space,
assuming we have $O(1)$ mutations at each site~\citep{kelleher2016efficient}.
For large sample sizes, this difference is profound, making it
conceivable, for example, to store the genetic ancestry
and variation data for the entire human population on a
laptop~\citep{kelleher2019inferring}. Besides the huge
difference in storage efficiency, it is also often far more efficient
to compute statistics of the sequence data from the
trees and mutations than it is to work with the sequences themselves.
For example, computing Tajima's $D$ from simulated data using \tskit\
is several orders of magnitude faster than efficient variant matrix
libraries for large sample sizes~\citep{ralph2020efficiently}.

@@

The software ecosystem built around \tskit\ is stable and mature,
with simulators such as
\fwdpy~\citep{thornton2014cpp},
\SLiM~\citep{haller2019slim},
\stdpopsim~\citep{adrion2020community},
\texttt{Geonomics}~\citep{de2021geonomics}
and
\texttt{GSpace}~\citep{virgoulay2021gspace},
and inference methods such as
\texttt{tsinfer}~\citep{kelleher2019inferring},
\texttt{tsdate}~\citep{,wohns2021unified}
and \texttt{Relate}~\citep{speidel2019method}
using either the Python or C
APIs to support outputting results in tree sequence format.
Tree sequences are stored in an efficient binary file format, and
are fully portable across operating systems and processor
architectures. The \tskit\ library ensures interoperability
between programs by having strict definitions of how the
information in each of the tables is interpreted, and
stringent checks for the internal consistency of the data model.

@@

The more recent trend has been to move away from this file and command-line
driven approach and to instead provide direct interfaces to the simulation
engines via an Application Programming Interface (API)~\citep[e.g.][]{
thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
The primary interface for \msprime\ is through a thoroughly documented and
stable Python
API, which has allowed a number of downstream tools to be built on top of
% Found these through the Dependent packages on GitHub
it~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
adrion2020community,adrion2020predicting, kamm2020efficiently,
mckenzie2020ipcoal, montinaro2020revisiting,
de2021geonomics,rivera2021simulation}.
As well as providing a stable and efficient platform for building
downstream applications, \msprime's Python API makes it much easier to
build reproducible simulation pipelines, as the entire workflow can
be encapsulated in a single script, and package and version
dependencies can be explicitly stated using the \texttt{pip}
or \texttt{conda} package managers.
For example, the errors made
in the influential simulation analysis of
\cite{martin2017human} were only detected because the pipeline
could be easily run and reanalysed~\citep{ragsdale2020lessons,martin2020erratum}.

A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
designed in part to avoid sources of error (see the Demography section) but
also to provide more appropriate defaults while keeping compatibility with
existing code. In the new APIs, ancestry and mutation simulation are fully
separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
functions replacing the legacy \texttt{simulate} function. Among other changes,
the new APIs default to discrete genome coordinates and finite sites mutations,
resolving a major source of confusion and error. The 0.x APIs are fully
supported and tested, and will be maintained for the foreseeable future.
We also provide an \ms\ compatible
command line interface to support existing workflows, and the
\texttt{msp} program that supports specifying demographic models in
Demes format~\citep{gower2021demes} from the command line, as
well as simulating mutations on a given tree sequence file.

% The output - what do we get out?
\subsection*{Data analysis}
\label{sec-data-analysis}
The point of running simulations is to generate data, and there
are two important ways in which the \tskit\ tree sequences output by
\msprime\ are more useful than established methods:
they are more efficient and they contain more information.
We will consider these points in turn.

@@

Classical text based output formats like \ms\ are inefficient to process,
but also lack a great deal of important information.
The tree-by-tree topology information output by simulators in Newick
format lacks any concept of node identity,
and means that we cannot reliably infer information about ancestors
from the output. Because Newick stores branch lengths rather
than node times, numerical precision issues also arise when summing
over many short branches~\citep{mcgill2013graphml}.
% JK: not sure this is worth leaving in tbh
This lack of reliable information about ancestors in simulation data
has lead, for example, to the definition of segments of
identity by descent as a set of samples having approximately the same time
to their most recent common ancestor along some section of the
genome~\citep{yang2016rapidly}.
Numerous forks (i.e., independent copies) of simulators have been created to access information not provided
in the output. For example, \ms\  has been forked to
output information about migrating segments~\citep{rosenzweig2016powerful},
information about ancestral lineages~\citep{chen2013asymptotic},
an array of population genetic statistics~\citep{ramos2007mlcoalsim}
% Any more?
and \ms's fork \texttt{msHOT}~\citep{hellenthal2007mshot},
has in turn been forked to output information on local
ancestry~\citep{racimo2017archaic}.
All of this information
is either directly available by default in \msprime, or can be optionally
stored via options such as \texttt{record\_migrations} or
\texttt{record\_full\_arg} (see the \nameref{sec-arg} section) and
can be efficiently and conveniently processed via \tskit\ APIs.

@@

Part of the reason for this poor record of software reuse and modularity is the
lack of standardised file formats, and in particular, the lack of common
library infrastructure to abstract the details of interchanging simulation
data. On the surface, it would appear that \msprime\ is also guilty of this
lack of modularity since it supports simulating both ancestry and mutations.
However, this is essentially a historical accident and because of the \tskit\
library, the mutation process in \msprime\ could just as easily
be performed in a separate package.
The mature and efficient C and Python interfaces for
\tskit\ makes it straightforward to add further information to
an existing file, and because of its efficient data interchange mechanisms,
there is no performance penalty for performing additional operations
in a different Python package.
Thanks to this interoperability, \msprime's
mutation generator can work with \emph{any} \tskit\ tree sequence,
be it simulated using \SLiM~\citep{haller2019slim} or
\fwdpy~\citep{thornton2014cpp}, or estimated from real
data~\citep{kelleher2019inferring,speidel2019method,wohns2021unified}.
It is a modular
component intended to fit into a larger software ecosystem, and
is in no way dependent on \msprime's ancestry simulator.

As well as providing a new API (see the \nameref{sec-sim-interface} section)
which emphasises the logical split between ancestry and mutation simulation,
we have greatly extended the sophistication of
\msprime's mutation generation engine for version 1.0,
achieving near feature-parity with \SeqGen.
We  support a large number of mutation models, including the
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and GTR~\citep{tavare1986some} nucleotide models
and the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
A general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
can be easily and efficiently defined in user code.
Mutation rates can vary along the genome, and multiple mutation
models can be imposed on a tree sequence by overlaying mutations
in multiple passes.
We have extensively validated the results of mutation simulations
against both theoretical expectations and output from
\SeqGen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}.

@@

Simulating mutations in \msprime\ is efficient.
Fig.~\ref{fig-mutations-perf} shows the time required to generate
mutations (using the default JC69 model) on
simulated tree sequences for a variety of mutation
rates as we vary the number of samples
(Fig.~\ref{fig-mutations-perf}A) and the sequence
length (Fig.~\ref{fig-mutations-perf}B).
For example, the longest running simulation in
Fig.~\ref{fig-mutations-perf}B required less than 2 seconds to
generate an average of 1.5 million mutations over 137,081 trees
in a tree sequence with 508,125 edges.
This efficiency for large numbers of trees is possible because
the tree sequence encoding allows us to generate mutations
on an edge-by-edge basis
(see Fig.~\ref{fig-ts-example} and the~\nameref{app-mutation-algorithm}
appendix),
rather than tree-by-tree and branch-by-branch as would otherwise be required.
In the above example from Fig.~\ref{fig-mutations-perf}B,
we would have to consider 273,887,838 branches
since there are 137,081 trees and 1,998 branches in each
tree ($\sim500$ times more work).
% python3 evaluation/generate_mutations_perf_data.py benchmark-single-tree
Even if we have a tree sequence consisting of a single tree
(so that there is no advantage to working edge-by-edge),
\msprime's mutation generator is still very efficient.
For example, we simulated mutations under the BLOSUM62 amino
acid model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim$260,000 mutations) in about $0.8$ seconds, including
the time required for file input and output.
We do not attempt a systematic benchmarking of \msprime's
mutation generation code against other methods because at this scale it is
difficult to disentangle the effects of inefficient input and
output formats from the mutation generation algorithms.
It seems quite unlikely, however,
that generating mutations with \msprime\ would be a bottleneck in any
realistic analysis.
